---
title: "Sample processig proficiency index"
author: "Taylor Johnson"
date: "2026-27-3"
output: html_document
runtime: shiny
---

# 1. Set up
```{r setup, include=FALSE}
source("config.R")
source("check.R")
source("plot_helpers.R")

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

pretty_group_label <- function(x) {
  gsub("_", " ", x, fixed = TRUE)
}
```

The validation pipeline covers:
✅ Required columns exist
✅ No missing values in required columns
✅ Required numeric columns contain numeric data
✅ No invalid negative values (Live ≥ 0, Dead ≥ 0)
✅ Positive values required for:
Isolation_event > 0
Total > 0
Tube_volume > 0

✅ Live ≤ Total
✅ Dead ≤ Total
✅ Live + Dead agrees with Total within the user-defined tolerance (COUNT_TOLERANCE = 2%)
✅ No duplicate records for the same:
Date
Parent_group
`r DEFAULT_COMPARISON_GROUP`
Isolation_event
Replicate

✅ Valid replicate structure:
Each isolation event contains exactly two technical replicates
Replicates must be labeled A and B
Each Date + Parent_group + `r DEFAULT_COMPARISON_GROUP` + Isolation_event combination contains exactly one A and one B

✅ Tube_volume normalization safety:
Tube_volume > Additive_vol
Blood_Volume remains positive after correction
Blood_Volume = Tube_volume - Additive_vol

✅ Row-specific error reporting for all validation failures

```{r}
# Load your reference dataset from the data directory.
reference_data <- read.csv(file.path(DATA_DIR, "example_reference_dataset.csv"))

# Validate and standardize the data.
reference_data <- validate_data(reference_data)

# Make sure the active comparison column exists after validation.
if (!DEFAULT_COMPARISON_GROUP %in% names(reference_data)) {
  stop(paste0("Comparison column not found after validation: ", DEFAULT_COMPARISON_GROUP))
}

comparison_group_values <- sort(unique(as.character(reference_data[[DEFAULT_COMPARISON_GROUP]])))
comparison_group_label <- DEFAULT_COMPARISON_GROUP

```

# 2. Calculate viability

```{r}
reference_data$Viability <- reference_data$Live / reference_data$Total
```

# 3. Normalize cell populations

```{r}
# Normalize cell counts per mL of estimated whole-blood volume.
reference_data$Total_norm <- reference_data$Total / reference_data$Blood_Volume
reference_data$Live_norm  <- reference_data$Live / reference_data$Blood_Volume
reference_data$Dead_norm  <- reference_data$Dead / reference_data$Blood_Volume
```

# 4. Calculate proportions

```{r}
# Proportion dead per total cell count.
reference_data$`Proportion dead` <- reference_data$Dead / reference_data$Total

# Proportion live per total cell count.
reference_data$`Proportion live` <- reference_data$Live / reference_data$Total
```

# 5. Calculate between-sample variability
## A. Format dataframe

```{r}
# Align replicate sample (i) cell counts and (ii) proportions in a single row
# for each unique combination of date, parent group, comparison group, and
# isolation event.
reference_data_wider <- reference_data |>
  dplyr::mutate(
    Sample_ID = as.character(Sample_ID),
    Date = parse_reference_date(Date)
  ) |>
  tidyr::pivot_wider(
    id_cols = c(
      Sample_ID,
      Date,
      Parent_group,
      dplyr::all_of(comparison_group_label),
      Isolation_event
    ),
    names_from = Replicate,
    values_from = c(Total, Total_norm, Live, Dead, Viability)
)

comparison_group_values <- sort(unique(as.character(reference_data_wider[[comparison_group_label]])))
```

## B. Variability calculation

```{r}
# Viability between replicate samples.
reference_data_wider$`Inter-sample Viability` <- 1 - abs(
  reference_data_wider$Viability_A - reference_data_wider$Viability_B
)

# Cell count agreement between replicate samples.
reference_data_wider$`Inter-sample Total Count` <- 1 - (
  abs(reference_data_wider$Total_A - reference_data_wider$Total_B) /
    ((reference_data_wider$Total_A + reference_data_wider$Total_B) / 2)
)
```

## C. Index calculation

```{r}
# Average calculated quality and consistency metrics for a summarized value
# indicating isolation efficacy across samples.
reference_data_wider$`Index` <- (
  (
    (reference_data_wider$Viability_A + reference_data_wider$Viability_B) / 2 +
      (reference_data_wider$`Inter-sample Viability` + reference_data_wider$`Inter-sample Total Count`) / 2
  ) / 2
)
```

## D. Parent-group summary

```{r}
Parent_group_summary <- reference_data_wider |>
  dplyr::group_by(Parent_group) |>
  dplyr::summarise(
    `Mean Index` = mean(`Index`, na.rm = TRUE),
    SD = sd(`Index`, na.rm = TRUE),
    .groups = "drop"
  )

Parent_group_longitudinal_summary <- reference_data_wider |>
  dplyr::group_by(Parent_group) |>
  dplyr::group_modify(~{
    .x <- dplyr::mutate(.x, Date = parse_reference_date(Date))

    model <- lm(`Index` ~ as.numeric(Date), data = .x)

    data.frame(
      `Annual Change (%)` = round(coef(model)[2] * 365.25 * 100, 1),
      `R-squared` = round(summary(model)$r.squared, 2)
    )
  })
```

# 6. Statistical analysis
## A. Dataset preparation

```{r}
comparison_group_list <- split(reference_data_wider, reference_data_wider[[comparison_group_label]])
comparison_group_values <- sort(unique(as.character(reference_data_wider[[comparison_group_label]])))
comparison_group_pairs <- if (length(comparison_group_values) >= 2) {
  combn(comparison_group_values, 2, simplify = FALSE)
} else {
  list()
}
```

## B. Statistical testing

```{r}
# Automated testing loop comparing two groups for each iteration.
results_list <- lapply(comparison_group_pairs, function(x) {
  comparison_1_name <- x[1]
  comparison_2_name <- x[2]
  df1 <- comparison_group_list[[comparison_1_name]]
  df2 <- comparison_group_list[[comparison_2_name]]

  safe_wilcox <- function(a, b) {
    tryCatch(
      wilcox.test(a, b, alternative = "two.sided"),
      error = function(e) NULL
    )
  }

  test <- safe_wilcox(df1$`Index`, df2$`Index`)
  viability_1 <- rowMeans(df1[, c("Viability_A", "Viability_B")], na.rm = TRUE)
  viability_2 <- rowMeans(df2[, c("Viability_A", "Viability_B")], na.rm = TRUE)
  viability_test <- safe_wilcox(viability_1, viability_2)
  inter_sample_viability_test <- safe_wilcox(df1$`Inter-sample Viability`, df2$`Inter-sample Viability`)
  inter_sample_total_test <- safe_wilcox(df1$`Inter-sample Total Count`, df2$`Inter-sample Total Count`)

  data.frame(
    Comparison = paste(comparison_1_name, "vs", comparison_2_name),
    metric = c("Index", "Mean Viability", "Inter-sample Viability", "Inter-sample Total"),
    stat = c(
      if (!is.null(test)) unname(test$statistic) else NA_real_,
      if (!is.null(viability_test)) unname(viability_test$statistic) else NA_real_,
      if (!is.null(inter_sample_viability_test)) unname(inter_sample_viability_test$statistic) else NA_real_,
      if (!is.null(inter_sample_total_test)) unname(inter_sample_total_test$statistic) else NA_real_
    ),
    p.value = c(
      if (!is.null(test)) test$p.value else NA_real_,
      if (!is.null(viability_test)) viability_test$p.value else NA_real_,
      if (!is.null(inter_sample_viability_test)) inter_sample_viability_test$p.value else NA_real_,
      if (!is.null(inter_sample_total_test)) inter_sample_total_test$p.value else NA_real_
    )
  )
})

if (length(results_list) > 0) {
  final_stats <- do.call(rbind, results_list)
  final_stats$p.adj <- p.adjust(final_stats$p.value, method = "BH")
} else {
  final_stats <- data.frame()
}

# Save reference data.
saveRDS(reference_data_wider, file.path(OUTPUT_DIR, "Index_data.rds"))
# Backward-compatible alias while you transition downstream scripts.
saveRDS(reference_data_wider, file.path(OUTPUT_DIR, "Index_data.rds"))

comparison_group_summary_levels <- comparison_group_values

comparison_group_summary_rows <- lapply(comparison_group_summary_levels, function(group_name) {
  
comparison_data <- reference_data_wider |>
  dplyr::filter(.data[[comparison_group_label]] == group_name) |>
  dplyr::mutate(Date = parse_reference_date(Date))

other_data <- reference_data_wider |>
  dplyr::filter(.data[[comparison_group_label]] != group_name)

  comparison_p <- tryCatch(
    wilcox.test(
      comparison_data$`Index`,
      other_data$`Index`
    )$p.value,
    error = function(e) NA_real_
  )

  annual_change <- NA_real_
  if (nrow(comparison_data) >= 2 && sum(!is.na(comparison_data$Date)) >= 2) {
    trend_model <- lm(`Index` ~ as.numeric(Date), data = comparison_data)
    annual_change <- round(coef(trend_model)[2] * 365.25 * 100, 1)
  }

  primary_driver <- "None"
  if (!is.na(comparison_p) && comparison_p < 0.05) {
    viability_p <- tryCatch(
      wilcox.test(
reference_data$Viability[reference_data[[comparison_group_label]] == group_name],
reference_data$Viability[reference_data[[comparison_group_label]] != group_name]
)$p.value,
      error = function(e) NA_real_
    )

    inter_viability_p <- tryCatch(
      wilcox.test(
        comparison_data$`Inter-sample Viability`,
        other_data$`Inter-sample Viability`
      )$p.value,
      error = function(e) NA_real_
    )

    inter_total_p <- tryCatch(
      wilcox.test(
        comparison_data$`Inter-sample Total Count`,
        other_data$`Inter-sample Total Count`
      )$p.value,
      error = function(e) NA_real_
    )

    component_p <- c(
      "Viability" = viability_p,
      "Inter-sample Viability" = inter_viability_p,
      "Inter-sample Total Count" = inter_total_p
    )

    component_p <- component_p[!is.na(component_p) & component_p < 0.05]

    if (length(component_p) > 0) {
      primary_driver <- paste(names(component_p), collapse = "; ")
    }
  }

  data.frame(
    comparison_group = group_name,
    `Mean Index` = round(mean(comparison_data$`Index`, na.rm = TRUE), 3),
    `SD` = round(sd(comparison_data$`Index`, na.rm = TRUE), 3),
    `Annual Change (%)` = annual_change,
    `comparison group vs repository p-value` = sprintf("%.3f", comparison_p),
    `Significant Variable Driving Difference` = primary_driver,
    check.names = FALSE
  )
})

comparison_group_summary <- dplyr::bind_rows(comparison_group_summary_rows)

# Repository-level summary.
repository_data <- reference_data_wider |>
  dplyr::mutate(Date = parse_reference_date(Date))

repository_model <- lm(
  `Index` ~ as.numeric(Date),
  data = repository_data
)

repository_summary <- data.frame(
  `Mean Index` = round(mean(repository_data$`Index`, na.rm = TRUE), 3),
  `SD` = round(sd(repository_data$`Index`, na.rm = TRUE), 3),
  `Number of Paired Isolations` = nrow(repository_data),
  `Annual Change (%)` = round(coef(repository_model)[2] * 365.25 * 100, 1),
  `R-squared` = round(summary(repository_model)$r.squared, 2),
  check.names = FALSE
)
```

# 7. Reference dataset - Shiny app visualization
## A. Define UI

```{r}
## A. Define UI

ui <- fluidPage(
  theme = app_theme,

  titlePanel(
    div(
      "Sample Processing Proficiency Index (SPPI) Dashboard",
      style = "font-weight:600; font-size:28px;"
    )
  ),

  br(),

  wellPanel(
    h4("Reference Dataset"),

    div(
      style = "max-width:300px;",
      fileInput(
        "reference_file",
        tags$strong("1. Upload CSV"),
        accept = ".csv",
        width = "100%"
      )
    ),

    div(
      style = "max-width:300px;",
      selectInput(
        "comparison_group",
        tags$strong("2. Compare by"),
        choices = setNames(
          c(
            "Technician",
            "Parent_group",
            "Cell_type",
            "Tube_type",
            "Protocol"
          ),
          pretty_group_label(
            c(
              "Technician",
              "Parent_group",
              "Cell_type",
              "Tube_type",
              "Protocol"
            )
          )
        ),
        selected = DEFAULT_COMPARISON_GROUP,
        width = "100%"
      )
    ),

    div(
      style = "max-width:300px;",
      tags$strong("3. Generate Reference Statistics"),
      br(),
      br(),
      actionButton(
        "build_reference",
        "Generate",
        class = "btn-primary",
        width = "100%"
      )
    ),

    br(),

    actionLink(
      "use_example_link",
      "Use example dataset"
    ),

    br(),
    br()
  ),

  navset_card_tab(
    nav_panel(
      "SPPI",

      wellPanel(
        h4("Interpretation"),
        p("The Sample Processing Proficiency Index (SPPI) incorporates isolation quality and consistency between replicate samples. Higher Indexs indicate better overall processing performance."),
      plotOutput("Index_plot", height = "70vh")
    )),

    nav_panel(
      "Longitudinal SPPI",

      navset_card_tab(
        nav_panel(
          "Comparison Group View",

          wellPanel(
            h4("Interpretation"),
            p("This view tracks Sample Processing Proficiency Index (SPPI) longitudinally for each selected comparison group."),
            tags$ul(
              tags$li("Positive annual change suggests improving sample processing performance."),
              tags$li("Negative annual change suggests declining sample processing performance."),
              tags$li("Lower p-values suggest the comparison group significantly differs from the repository.")
            )
          ),

          plotOutput("time_plot", height = "60vh"),
          br(),
          tableOutput("comparison_group_summary")
        ),

        nav_panel(
          "Repository View",

          wellPanel(
            h4("Interpretation"),
            p("This view tracks Sample Processing Proficiency Index (SPPI) longitudinally at the repository level, independent of the comparison group."),
            tags$ul(
              tags$li("Positive annual change suggests improving sample processing performance."),
              tags$li("Negative annual change suggests declining sample processing performance.")
            )
          ),

          plotOutput("repository_time_plot", height = "60vh"),
          br(),
          tableOutput("repository_summary")
        )
      )
    ),

    nav_panel(
      "Component Metrics",

      navset_card_tab(
        nav_panel(
          "Viability",

          wellPanel(
            h4("Interpretation"),
            p("The Sample Processing Proficiency Index (SPPI) quality component is captured by viability - the proportion of recovered, live cells.")
          ),

          plotOutput("viability_plot", height = "90vh")
        ),

        nav_panel(
          "Inter-sample Variability",

          wellPanel(
            h4("Interpretation"),
            p("The Sample Processing Proficiency Index (SPPI) consistency component is captured by an inter-tube viability and total cell count composite."),
            tags$ul(
              tags$li("Higher values indicate better metric agreement between replicate samples."),
              tags$li("Lower values suggest increased variability between replicate samples during processing.")
            )
          ),

          selectInput(
            "inter_type",
            "Variability Metric",
            choices = c(
              "Viability" = "Inter-sample Viability",
              "Total Count" = "Inter-sample Total Count"
            ),
            selected = "Inter-sample Viability"
          ),

          plotOutput("inter_variability_combined_plot", height = "70vh")
        )
      )
    ),

    nav_panel(
      "Benchmarking Assessment",

      fluidRow(
        column(
          5,
          wellPanel(
            h4("Inputs"),

            numericInput(
              "new_tube_volume_A",
              "Tube A tube volume (mL)",
              value = 0,
              min = 0,
              step = 0.1
            ),
            numericInput(
              "new_tube_volume_B",
              "Tube B tube volume (mL)",
              value = 0,
              min = 0,
              step = 0.1
            ),
            numericInput(
              "new_additive_vol",
              "Additive vol (mL)",
              value = 0,
              min = 0,
              step = 0.1
            ),
            numericInput(
              "new_total_A",
              "Tube A total count",
              value = 0,
              min = 0,
              step = 1
            ),
            numericInput(
              "new_total_B",
              "Tube B total count",
              value = 0,
              min = 0,
              step = 1
            ),
            numericInput(
              "new_live_A",
              "Tube A live count",
              value = 0,
              min = 0,
              step = 1
            ),
            numericInput(
              "new_live_B",
              "Tube B live count",
              value = 0,
              min = 0,
              step = 1
            ),

            actionButton(
              "go_benchmarking_assessment",
              "Calculate Benchmarking Assessment SPPI",
              class = "btn-primary",
              width = "100%"
            )
          )
        ),

        column(
          7,
          wellPanel(
            h4("Interpretation"),
            p("This Sample Processing Proficiency Index (SPPI) benchmark compares a single paired isolation event to the loaded reference dataset Higher SPPI indicates better overall processing performance.")
          ),

          plotOutput("benchmarking_assessment_Index_plot", height = "48vh"),
          br(),
          
fluidRow(
  column(
    width = 6,
    wellPanel(
      h4("Loaded Reference Statistics"),
      tableOutput("benchmarking_reference")
    )
  ),
  column(
    width = 6,
    wellPanel(
      h4("Benchmarked Statistics"),
      tableOutput("benchmarking_assessment_summary")
    )
  )
),          
          br(),
          
          textOutput("benchmarking_assessment_Index_note")
        )
      )
    )
  )
)

```

## B. Define server
```{r}

server <- function(input, output, session) {

  data_source_mode <- reactiveVal("upload")

  observeEvent(input$comparison_group, {
    DEFAULT_COMPARISON_GROUP <<- input$comparison_group
  }, ignoreInit = FALSE)

  active_group_label <- reactive({
    req(input$comparison_group)
    input$comparison_group
  })

  reference_raw <- reactive({
    if (identical(data_source_mode(), "example")) {
      read.csv(
        file.path(DATA_DIR, "example_reference_dataset.csv"),
        stringsAsFactors = FALSE,
        check.names = FALSE
      )
    } else {
      req(input$reference_file)
      read.csv(
        input$reference_file$datapath,
        stringsAsFactors = FALSE,
        check.names = FALSE
      )
    }
  })

  reference_data <- reactive({
    df <- reference_raw()
    df <- validate_data(df)

    df$Viability <- df$Live / df$Total
    df$Total_norm <- df$Total / df$Blood_Volume
    df$Live_norm  <- df$Live / df$Blood_Volume
    df$Dead_norm  <- df$Dead / df$Blood_Volume
    df$`Proportion dead` <- df$Dead / df$Total
    df$`Proportion live` <- df$Live / df$Total

    df
  })

  reference_data_wider <- reactive({
    df <- reference_data()
    group_col <- active_group_label()

    wider <- df |>
      dplyr::mutate(
        Date = parse_reference_date(Date)
      ) |>
      tidyr::pivot_wider(
        id_cols = c(
          Date,
          Parent_group,
          dplyr::all_of(group_col),
          Isolation_event
        ),
        names_from = Replicate,
        values_from = c(Total, Total_norm, Live, Dead, Viability)
      )

    wider$`Inter-sample Viability` <- 1 - abs(
      wider$Viability_A - wider$Viability_B
    )

    wider$`Inter-sample Total Count` <- 1 - (
      abs(wider$Total_norm_A - wider$Total_norm_B) /
        ((wider$Total_norm_A + wider$Total_norm_B) / 2)
    )

    wider$Index <- (
      (
        (wider$Viability_A + wider$Viability_B) / 2 +
          (
            wider$`Inter-sample Viability` +
              wider$`Inter-sample Total Count`
          ) / 2
      ) / 2
    )

    wider
  })

  comparison_group_values <- reactive({
    group_col <- active_group_label()
    vals <- unique(as.character(reference_data_wider()[[group_col]]))
    sort(vals)
  })

  comparison_group_pairs <- reactive({
    vals <- comparison_group_values()
    if (length(vals) >= 2) {
      combn(vals, 2, simplify = FALSE)
    } else {
      list()
    }
  })

  comparison_group_summary <- reactive({
    df <- reference_data_wider() |>
      dplyr::mutate(Date = parse_reference_date(Date))

    group_col <- active_group_label()
    groups <- sort(unique(as.character(df[[group_col]])))

    if (length(groups) == 0) {
      return(data.frame())
    }

    rows <- lapply(groups, function(group_name) {
      group_data <- df |>
        dplyr::filter(.data[[group_col]] == group_name)

      repo_data <- df |>
        dplyr::filter(.data[[group_col]] != group_name)

      comparison_p <- tryCatch(
        wilcox.test(group_data$Index, repo_data$Index)$p.value,
        error = function(e) NA_real_
      )

      annual_change <- NA_real_
      if (nrow(group_data) >= 2 && sum(!is.na(group_data$Date)) >= 2) {
        trend_model <- lm(Index ~ as.numeric(Date), data = group_data)
        annual_change <- round(coef(trend_model)[2] * 365.25 * 100, 1)
      }

      primary_driver <- "None"

      if (!is.na(comparison_p) && comparison_p < 0.05) {
        viability_p <- tryCatch(
          wilcox.test(group_data$Viability_A, repo_data$Viability_A)$p.value,
          error = function(e) NA_real_
        )

        inter_viability_p <- tryCatch(
          wilcox.test(
            group_data$`Inter-sample Viability`,
            repo_data$`Inter-sample Viability`
          )$p.value,
          error = function(e) NA_real_
        )

        inter_total_p <- tryCatch(
          wilcox.test(
            group_data$`Inter-sample Total Count`,
            repo_data$`Inter-sample Total Count`
          )$p.value,
          error = function(e) NA_real_
        )

        sig_components <- c(
          "Viability" = viability_p,
          "Inter-sample Viability" = inter_viability_p,
          "Inter-sample Total Count" = inter_total_p
        )

        sig_components <- sig_components[
          !is.na(sig_components) & sig_components < 0.05
        ]

        if (length(sig_components) > 0) {
          primary_driver <- paste(names(sig_components), collapse = "; ")
        }
      }

data.frame(
  `Comparison group` = group_name,
  `N` = nrow(group_data),
  `Mean Index` = round(mean(group_data$Index, na.rm = TRUE), 3),
  `SD` = round(sd(group_data$Index, na.rm = TRUE), 3),
  `Annual Change (%)` = annual_change,
  `Comparison group vs repository p-value` = sprintf("%.3f", comparison_p),
  `Significant Variable Driving Difference` = primary_driver,
  check.names = FALSE
)
          })

    dplyr::bind_rows(rows)
  })

  repository_summary <- reactive({
    df <- reference_data_wider() |>
      dplyr::mutate(Date = parse_reference_date(Date))

    repository_model <- tryCatch(
      lm(Index ~ as.numeric(Date), data = df),
      error = function(e) NULL
    )

    annual_change <- NA_real_
    r_squared <- NA_real_

    if (!is.null(repository_model)) {
      annual_change <- round(coef(repository_model)[2] * 365.25 * 100, 1)
      r_squared <- round(summary(repository_model)$r.squared, 2)
    }

    data.frame(
      `Mean Index` = round(mean(df$Index, na.rm = TRUE), 3),
      `SD` = round(sd(df$Index, na.rm = TRUE), 3),
      `Number of Paired Isolations` = nrow(df),
      `Annual Change (%)` = annual_change,
      `R-squared` = r_squared,
      check.names = FALSE
    )
  })

  reference_statistics <- reactive({
    list(
      Index = list(
        mean = mean(reference_data_wider()$Index, na.rm = TRUE),
        sd = sd(reference_data_wider()$Index, na.rm = TRUE),
        median = median(reference_data_wider()$Index, na.rm = TRUE),
        q25 = unname(quantile(reference_data_wider()$Index, 0.25, na.rm = TRUE)),
        q75 = unname(quantile(reference_data_wider()$Index, 0.75, na.rm = TRUE))
      ),
      viability = list(
        mean = mean(reference_data()$Viability, na.rm = TRUE),
        sd = sd(reference_data()$Viability, na.rm = TRUE),
        median = median(reference_data()$Viability, na.rm = TRUE)
      ),
      inter_sample_viability = list(
        mean = mean(reference_data_wider()$`Inter-sample Viability`, na.rm = TRUE),
        sd = sd(reference_data_wider()$`Inter-sample Viability`, na.rm = TRUE)
      ),
      inter_sample_total = list(
        mean = mean(reference_data_wider()$`Inter-sample Total Count`, na.rm = TRUE),
        sd = sd(reference_data_wider()$`Inter-sample Total Count`, na.rm = TRUE)
      ),
      comparison_group_summary = comparison_group_summary(),
      repository_summary = repository_summary()
    )
  })

  reference_source_label <- reactive({
    if (identical(data_source_mode(), "example")) {
      "default example reference dataset"
    } else {
      "user-uploaded reference dataset"
    }
  })

  observeEvent(input$use_example_link, {
    data_source_mode("example")
  })

  observeEvent(input$reference_file, {
    data_source_mode("upload")
  })

benchmarking_assessment_metrics <- eventReactive(input$go_benchmarking_assessment, {

  validate(
    need(
      input$new_total_A > 0 && input$new_total_B > 0,
      "Total counts must be greater than zero."
    ),
    need(
      input$new_live_A >= 0 && input$new_live_B >= 0,
      "Live counts cannot be negative."
    ),
    need(
      input$new_live_A <= input$new_total_A &&
        input$new_live_B <= input$new_total_B,
      "Live count cannot exceed total count."
    ),
    need(
      input$new_tube_volume_A > input$new_additive_vol &&
        input$new_tube_volume_B > input$new_additive_vol,
      "Tube volume must be greater than Additive vol."
    )
  )

  blood_vol_A <- as.numeric(input$new_tube_volume_A - input$new_additive_vol)
  blood_vol_B <- as.numeric(input$new_tube_volume_B - input$new_additive_vol)

  viab_A <- as.numeric(input$new_live_A / input$new_total_A)
  viab_B <- as.numeric(input$new_live_B / input$new_total_B)

  total_norm_A <- as.numeric(input$new_total_A / blood_vol_A)
  total_norm_B <- as.numeric(input$new_total_B / blood_vol_B)

  inter_viab <- as.numeric(1 - abs(viab_A - viab_B))
  inter_total <- as.numeric(
    1 - abs(total_norm_A - total_norm_B) / mean(c(total_norm_A, total_norm_B))
  )

  sppi <- as.numeric(
    (((viab_A + viab_B) / 2) + ((inter_viab + inter_total) / 2)) / 2
  )

  list(
    tube_volume_A = as.numeric(input$new_tube_volume_A),
    tube_volume_B = as.numeric(input$new_tube_volume_B),
    additive_vol = as.numeric(input$new_additive_vol),
    blood_volume_A = blood_vol_A,
    blood_volume_B = blood_vol_B,
    total_A = as.numeric(input$new_total_A),
    total_B = as.numeric(input$new_total_B),
    live_A = as.numeric(input$new_live_A),
    live_B = as.numeric(input$new_live_B),
    viability_A = viab_A,
    viability_B = viab_B,
    total_norm_A = total_norm_A,
    total_norm_B = total_norm_B,
    inter_viability = inter_viab,
    inter_total = inter_total,
    Index = sppi
  )
})

benchmarking_assessment_reference_stats <- reactive({
  Indexs <- as.numeric(reference_data_wider()$Index)

  list(
    mean = mean(Indexs, na.rm = TRUE),
    sd = sd(Indexs, na.rm = TRUE),
    median = median(Indexs, na.rm = TRUE),
    q25 = unname(quantile(Indexs, 0.25, na.rm = TRUE)),
    q75 = unname(quantile(Indexs, 0.75, na.rm = TRUE))
  )
})

benchmarking_assessment_percentile <- reactive({
  df <- reference_data_wider()
  user_Index <- as.numeric(benchmarking_assessment_metrics()$Index)
  as.numeric(mean(as.numeric(df$Index) <= user_Index, na.rm = TRUE) * 100)
})



output$benchmarking_assessment_Index_note <- renderText({
  paste0(
    "Benchmark source: ", reference_source_label(), ". "
  )
})

  benchmarking_assessment_reference_stats <- reactive({
    Indexs <- reference_data_wider()$Index

    list(
      mean = mean(Indexs, na.rm = TRUE),
      sd = sd(Indexs, na.rm = TRUE),
      median = median(Indexs, na.rm = TRUE),
      q25 = quantile(Indexs, 0.25, na.rm = TRUE),
      q75 = quantile(Indexs, 0.75, na.rm = TRUE)
    )
  })

  benchmarking_assessment_percentile <- reactive({
    df <- reference_data_wider()
    Index <- benchmarking_assessment_metrics()$Index
    mean(df$Index <= Index, na.rm = TRUE) * 100
  })

  output$benchmarking_assessment_Index_plot <- renderPlot({
    metrics <- benchmarking_assessment_metrics()
    req(metrics)

    df <- reference_data_wider()
    req(nrow(df) > 0)

    user_Index <- metrics$Index
    user_pct <- benchmarking_assessment_percentile()
    ref_stats <- benchmarking_assessment_reference_stats()

    ggplot2::ggplot(df, ggplot2::aes(x = "", y = Index)) +
      ggplot2::geom_violin(
        fill = COLORS$primary,
        alpha = 0.20,
        color = "white",
        linewidth = 0.8
      ) +
      ggplot2::geom_jitter(
        width = 0.08,
        alpha = 0.50,
        size = 2.5,
        color = "white"
      ) +
      ggplot2::geom_hline(
        yintercept = user_Index,
        linetype = "dashed",
        linewidth = 1.1,
        color = COLORS$accent_red
      ) +
      ggplot2::annotate(
        "label",
        x = 1,
        y = user_Index,
        label = paste0("Benchmarked SPPI = ", round(user_Index, 3)),
        fill = COLORS$accent_red,
        color = "white",
        fontface = "bold",
        size = 5
      ) +
      ggplot2::labs(
        title = "SPPI Benchmarking Assessment",
        subtitle = paste0(
          "Reference source: ", reference_source_label(), " | ",
          "Percentile vs reference: ", round(user_pct, 1), "% | ",
          "Reference mean: ", round(ref_stats$mean, 3)
        ),
        x = NULL,
        y = "SPPI"
      ) +
      custom_theme +
      ggplot2::theme(
        axis.text.x = ggplot2::element_blank(),
        axis.ticks.x = ggplot2::element_blank()
      )
  }, bg = COLORS$bg_dark)
```

```{r}

output$benchmarking_reference <- renderTable({
  metrics <- benchmarking_assessment_metrics()
  req(metrics)

  stats <- benchmarking_assessment_reference_stats()

  data.frame(
    Metric = c(
      "Reference SPPI mean",
      "Reference SPPI SD",
      "Reference SPPI median"
    ),
    Value = c(
      sprintf("%.3f", stats$mean),
      sprintf("%.3f", stats$sd),
      sprintf("%.3f", stats$median)
    ),
    check.names = FALSE
  )
}, striped = TRUE, bordered = TRUE, spacing = "s")


  output$benchmarking_assessment_summary <- renderTable({
    metrics <- benchmarking_assessment_metrics()
    req(metrics)
    stats <- benchmarking_assessment_reference_stats()
    user_pct <- benchmarking_assessment_percentile()
    data.frame(
      Metric = c(
        "Benchmarked SPPI",
        "Benchmarked percentile",
        "Tube A viability",
        "Tube B viability",
        "Inter-sample viability",
        "Inter-sample total agreement"
      ),
      Value = c(
        sprintf("%.3f", metrics$Index),
        sprintf("%.1f%%", user_pct),
        sprintf("%.3f", metrics$viability_A),
        sprintf("%.3f", metrics$viability_B),
        sprintf("%.3f", metrics$inter_viability),
        sprintf("%.3f", metrics$inter_total)
      ),
      check.names = FALSE
    )
  }, striped = TRUE, bordered = TRUE, spacing = "s")


```

```{r}
  output$benchmarking_assessment_Index_note <- renderText({
    paste0(
      "Benchmark source: ", reference_source_label(), ". "
    )
  })

  observeEvent(input$build_reference, {
    req(reference_data_wider())

    saveRDS(
      reference_data_wider(),
      file = file.path(OUTPUT_DIR, "reference_index_data.rds")
    )

    saveRDS(
      reference_statistics(),
      file = file.path(OUTPUT_DIR, "reference_statistics.rds")
    )

    showNotification(
      "Reference Rds files saved to the output directory.",
      type = "message"
    )
  })

  output$Index_plot <- renderPlot({
    req(length(comparison_group_values()) >= 2)

    p <- betweenstats(
      data = reference_data_wider(),
      x = active_group_label(),
      y = "Index",
      type = "nonparametric",
      results.subtitle = TRUE,
      display.stats = TRUE
    ) +
      ggplot2::labs(
        x = pretty_group_label(active_group_label()),
        title = paste("SPPI by", pretty_group_label(active_group_label())),
        subtitle = "Each point represents one paired isolation event"
      )

    if (length(comparison_group_values()) == 2) {
      p_value <- wilcox.test(
        reference_data_wider()$Index[
          reference_data_wider()[[active_group_label()]] == comparison_group_values()[1]
        ],
        reference_data_wider()$Index[
          reference_data_wider()[[active_group_label()]] == comparison_group_values()[2]
        ]
      )$p.value

      p <- p +
        ggsignif::geom_signif(
          comparisons = list(comparison_group_values()),
          annotations = format_p_label(p_value),
          color = COLORS$primary,
          textsize = 5
        )
    }

    p 
    }, bg = "#222222")

  output$viability_plot <- renderPlot({
    req(length(comparison_group_values()) >= 2)

    p <- betweenstats(
      data = reference_data(),
      x = active_group_label(),
      y = "Viability",
      type = "nonparametric",
      results.subtitle = TRUE,
      display.stats = TRUE
    ) +
      ggplot2::labs(
        x = pretty_group_label(active_group_label()),
        title = paste("Viability by", pretty_group_label(active_group_label())),
        subtitle = "Each point represents one paired isolation event"
      )

    if (length(comparison_group_values()) == 2) {
      p_value <- wilcox.test(
        reference_data()$Viability[
          reference_data()[[active_group_label()]] == comparison_group_values()[1]
        ],
        reference_data()$Viability[
          reference_data()[[active_group_label()]] == comparison_group_values()[2]
        ]
      )$p.value

      p <- p +
        ggsignif::geom_signif(
          comparisons = list(comparison_group_values()),
          annotations = format_p_label(p_value),
          color = COLORS$primary,
          textsize = 5
        )
    }

    p 
  }, bg = "#222222")

  output$inter_variability_combined_plot <- renderPlot({
    req(input$inter_type)
    req(length(comparison_group_values()) >= 2)

    p <- betweenstats(
      data = reference_data_wider(),
      x = active_group_label(),
      y = input$inter_type,
      type = "nonparametric",
      results.subtitle = TRUE,
      display.stats = TRUE,
      y_label = input$inter_type
    ) +
      ggplot2::labs(
        x = pretty_group_label(active_group_label()),
        title = paste("SPPI by", pretty_group_label(active_group_label())),
        subtitle = "Each point represents one paired isolation event"
      )

    if (length(comparison_group_values()) == 2) {
      metric_vals <- reference_data_wider()[[input$inter_type]]

      p_value <- wilcox.test(
        metric_vals[
          reference_data_wider()[[active_group_label()]] == comparison_group_values()[1]
        ],
        metric_vals[
          reference_data_wider()[[active_group_label()]] == comparison_group_values()[2]
        ]
      )$p.value

      p <- p +
        ggsignif::geom_signif(
          comparisons = list(comparison_group_values()),
          annotations = format_p_label(p_value),
          color = COLORS$primary,
          textsize = 5
        )
    }

    p
  }, bg = "#222222")

  output$time_plot <- renderPlot({
    time_data <- reference_data_wider() |>
      dplyr::mutate(Date = parse_reference_date(Date)) |>
      dplyr::filter(
        !is.na(Date),
        !is.na(.data[["Index"]])
      )

    ggplot2::ggplot(
      time_data,
      ggplot2::aes(
        x = Date,
        y = .data[["Index"]],
        color = .data[[active_group_label()]]
      )
    ) +
      ggplot2::geom_point(size = 3, alpha = 0.8) +
      ggplot2::geom_smooth(
        ggplot2::aes(group = .data[[active_group_label()]]),
        method = "lm",
        se = FALSE,
        linewidth = 0.9
      ) +
      ggplot2::scale_y_continuous(
        limits = c(0.75, 1),
        breaks = seq(0.75, 1, by = 0.05)
      ) +
      adaptive_date_scale(time_data$Date) +
      ggplot2::labs(
        title = paste("Longitudinal SPPI by", pretty_group_label(active_group_label())),
        subtitle = paste0(
          "Each point represents one paired isolation event; lines show ",
          active_group_label(),
          "-level linear trends"
        ),
        x = "Date",
        y = "Index",
        color = active_group_label()
      ) +
      custom_theme +
      ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
      )
  }, bg = "#222222")

  output$repository_time_plot <- renderPlot({
    repository_data <- reference_data_wider() |>
      dplyr::mutate(Date = parse_reference_date(Date))

    ggplot2::ggplot(
      repository_data,
      ggplot2::aes(
        x = Date,
        y = Index
      )
    ) +
      ggplot2::geom_point(
        size = 3,
        alpha = 0.8,
        color = COLORS$primary
      ) +
      ggplot2::geom_smooth(
        method = "lm",
        se = FALSE,
        linewidth = 0.9,
        color = "white"
      ) +
      ggplot2::scale_y_continuous(
        limits = c(0.75, 1),
        breaks = seq(0.75, 1, by = 0.05)
      ) +
      adaptive_date_scale(repository_data$Date) +
      ggplot2::labs(
        title = paste("Longitudinal repository SPPI by", pretty_group_label(active_group_label())),
        subtitle = "Each point represents one paired isolation event; lines show repository-level linear trends",
        x = "Date",
        y = "Index"
      ) +
      custom_theme +
      ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
      )
  }, bg = "#222222")

  output$comparison_group_summary <- renderTable({
    comparison_group_summary()
  }, striped = TRUE, bordered = TRUE, spacing = "s")

  output$repository_summary <- renderTable({
    repository_summary()
  }, striped = TRUE, bordered = TRUE, spacing = "s")
}

```

# 8. Save reference statistics

```{r}
reference_statistics <- list(

  Index = list(
    mean = mean(reference_data_wider$`Index`, na.rm = TRUE),
    sd = sd(reference_data_wider$`Index`, na.rm = TRUE),
    median = median(reference_data_wider$`Index`, na.rm = TRUE),
    q25 = quantile(reference_data_wider$`Index`, 0.25, na.rm = TRUE),
    q75 = quantile(reference_data_wider$`Index`, 0.75, na.rm = TRUE)
  ),

  viability = list(
    mean = mean(reference_data$Viability, na.rm = TRUE),
    sd = sd(reference_data$Viability, na.rm = TRUE),
    median = median(reference_data$Viability, na.rm = TRUE)
  ),

  inter_sample_viability = list(
    mean = mean(reference_data_wider$`Inter-sample Viability`, na.rm = TRUE),
    sd = sd(reference_data_wider$`Inter-sample Viability`, na.rm = TRUE)
  ),

  inter_sample_total = list(
    mean = mean(reference_data_wider$`Inter-sample Total Count`, na.rm = TRUE),
    sd = sd(reference_data_wider$`Inter-sample Total Count`, na.rm = TRUE)
  )

)

saveRDS(
  reference_statistics,
  file.path(
    OUTPUT_DIR,
    "reference_statistics.rds"
  )
)

saveRDS(
  comparison_group_summary,
  file.path(
    OUTPUT_DIR,
    "comparison_group_summary.rds"
  )
)
```

# 9. Run Shiny App

```{r}
# Run user interface and server with ShinyApp.
shiny::shinyApp(ui = ui, server = server)
```

